Unshuffling Permutations

### 2015-09-16 article.tex
nouvelle version avec les preuves qui manquaient
Debut homogeneisation

For any nonnegative integer $n$, $[n]$ is the set $\{1, \dots, n\}$.
We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. The {\em empty word} is
denoted by $\epsilon$. If $u$ is a word, we denote by $\widetilde{u}$
the {\em mirror image} of $u$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$. The {\em shuffle} of two words $u$ and
$v$ is the set recursively defined by
$u \shuffle \epsilon = u = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a {\em square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. If $\pi$ is a permutation, we denote
by $\bar \pi$ (resp.~$\pi^{-1}$) the {\em complementary} (resp.
{\em inverse}) of $\pi$. If $u$ is a word of integers without multiple
occurrences of a same letter, $\STD(u)$ is the {\em standarized} of $u$,
that is the unique permutation of the same size as $u$ such that for all
$i, j \in [|u|]$, $u_i < u_j$ if and only if $\STD(u)_i < \STD(u)_j$. In
particular, the image of the map $\STD$ is the set $S$ of all
permutations. Two words $u$ and $v$ having the same standarized are
{\em order-isomorphic}. If $\sigma$ is a permutation, there is an
{\em occurrence} of (the {\em pattern}) $\sigma$ in $\pi$ if there is a
set $P$ of indexes of letters of $\pi$ such that $\sigma$ and $\pi_{|P}$
are order-isomorphic. When $\pi$ does not admit any occurrence of
$\sigma$, $\pi$ {\em avoids} $\sigma$.
We write $\nearrow_{n}$ (resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
we write seqeunce of integers $s = s_1 s_2 \ldots s_n$, we
we let $s\;[k]$ stand for the lifted sequence
$(s_1+k) s_2+k) \ldots (s_n+k)$.

@@

    Next, since by definition of the complementation operation on
    permutations, for any permutation $\tau$ and any indexes $i$ and $k$,
    we have $\tau_i < \tau_k$ if and only if $\bar \tau_i > \bar \tau_k$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_2(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_2(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_2(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\bar \pi_{|P_1}\right)
        \otimes \STD\left(\bar \pi_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_2\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_2\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_2 \otimes \phi_2) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_2$ is a morphism of coalgebras and hence, that
    $\phi_2$ is a morphism of associative algebras.
    \smallskip

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau_i : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection, and since
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
    \qed
\end{proof}
\medskip

@@

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item
    $\pi$ is a square.
    \item
    There exists a perfect containment-free mathing
    $\mathcal{M}$ of $G_\pi$ such that
    for any two disctinct edges
    $(a, b)$ and $(a', b')$ in $\mathcal{M}$
    we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

  (\emph{1} $\Rightarrow$ \emph{2}).
  Let $\sigma$ be a permutation such that $\pi$ is the union of two
  patterns that are order-isomorphic to $\sigma$.
  Let $2n = |\pi|$ (and hence $|\sigma| = n$).
  Fix the occurrences in $\pi$ of the two disjoint patterns that
  are order-isomorphic to $\sigma$ and
  write
  $I^1 = \{i^1_1, i^1_2, \ldots, i^1_n\}$,
  $i^1_1 < i^1_2 < \ldots < i^1_n$, for the positions in $\pi$
  of the first pattern order-ismorphic to $\sigma$ and
  $I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$,
  $i^2_1 < i^2_2 < \ldots < i^2_n$, for the positions in $\pi$
  of the second pattern order-ismorphic to $\sigma$.
  Let $G_\pi$ the linear clique associated to $\pi$ with
  vertices $\VS{G_\pi} =
  \{i^1_j : 1 \leq j \leq n\} \cup \{i^2_j : 1 \leq j \leq n\}$.
  It is easily seen that
  $\mathcal{M} = \{\{i^1_j, i^2_j\} : 1 \leq j \leq n\}$
  is a subset of $\ES{G_\pi}$.
  Furthermore, $\mathcal{M}$ is a perfect matching since
  $I^1 \cap I^2 = \emptyset$,
  and $I^1 \cup I^2 = \{1, 2,  \ldots 2n\}$.
  It is also containment-free.
  Indeed, if it were not the case then there would exist two edges
  $e = \{i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$, $j < k$,
  in $\mathcal{M}$ such that
  $i^1_j < i^1_k < i^2_k < i^2_j$ or
  $i^1_j < i^2_k < i^1_k < i^2_j$.
  This is a contradiction since
  $i^2_k > i^2_j$ if $k > j$.
  Finally, for any two distinct edges
  $e = \{\i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
  we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
  $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing
  in both cases two elements (at position $j$ and $k$)
  in two patterns that are order-isomorphic to $\sigma$.

@@

\begin{proof}
  The problem is certainly in $\NPclass$.
  We propose a reduction from the pattern involvement problem which is
  known to be \NPC~\cite{Bose:Buss:Lubiw:1998}:
  Given permutation $\pi$ and $\sigma$, decide whether $\sigma$
  occurs in $\pi$ (as an order-isomorphic pattern).

@@

      \node (claim edge 1) at (8.5,8.5) {Claim~\ref{}};

@@

      \node (claim edge 1) at (6.5,4) {Claim~\ref{}};

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, how many binary
strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?)
% Given two
% permutations $\pi$ and $\sigma$, how hard is the problem of deciding
% whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip

>>>>

If $S$ is a finite set, the cardinality of $S$ is denoted by $|S|$,
and if $P$ and $Q$ are two disjoint sets, $P \sqcup Q$ denotes the
disjoint union of $P$ and $Q$. For any nonnegative integer $n$, $[n]$
is the set $\{1, \dots, n\}$.

We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. Let $u$ be a word. The
length of $u$ is denoted by $|u|$. The {\em empty word}, the only word
of null length, is denoted by $\epsilon$. We denote by $\widetilde{u}$
the {\em mirror image} of $u$ that is the word
$u_{|u|} u_{|u| - 1} \dots u_1$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$.
If $u$ is a word of integers and $k$ is an integer, we denote by
$u[k]$ the word obtained by incrementing by $k$ all letters of $u$.
The {\em shuffle} of two words $u$ and
$v$ is the set recursively defined by
$u \shuffle \epsilon = u = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a {\em square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations.
For any nonnegative integer $n$, we write $\nearrow_{n}$
(resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
If $\pi$ is a permutation of $S_n$,
we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is the
permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

If $u$ is a word of integers without multiple
occurrences of a same letter, $\STD(u)$ is the {\em standarized} of $u$,
that is the unique permutation of the same size as $u$ such that for all
$i, j \in [|u|]$, $u_i < u_j$ if and only if $\STD(u)_i < \STD(u)_j$. In
particular, the image of the map $\STD$ is the set $S$ of all
permutations. Two words $u$ and $v$ having the same standarized are
{\em order-isomorphic}. If $\sigma$ is a permutation, there is an
{\em occurrence} of (the {\em pattern}) $\sigma$ in $\pi$ if there is a
set $P$ of indexes of letters of $\pi$ such that $\sigma$ and $\pi_{|P}$
are order-isomorphic. When $\pi$ does not admit any occurrence of
$\sigma$, $\pi$ {\em avoids} $\sigma$.

@@

    Next, since by definition of the complementation operation on
    permutations, for any permutation $\tau$ and any indexes $i$ and $k$,
    we have $\tau(i) < \tau(k)$ if and only if $\bar \tau(i) > \bar \tau(k)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_2(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_2(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_2(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\bar \pi_{|P_1}\right)
        \otimes \STD\left(\bar \pi_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_2\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_2\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_2 \otimes \phi_2) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_2$ is a morphism of coalgebras and hence, that
    $\phi_2$ is a morphism of associative algebras.
    \smallskip

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau(i) : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection, and since
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
    \qed
\end{proof}
\medskip

@@

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item
    The permutation $\pi$ is a square.
    \item
    There exists a perfect containment-free mathing
    $\mathcal{M}$ of $G_\pi$ such that
    for any two disctinct edges
    $(a, b)$ and $(a', b')$ in $\mathcal{M}$
    we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

  (\emph{1} $\Rightarrow$ \emph{2}).
  Since $\pi$ is a quare, it admits a square root $\sigma$. Then, by
  definition, there
  are two sets $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$ and
  $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
  letters of $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
  order-isomorphic to $\sigma$.
  %Let $\sigma$ be a permutation such that $\pi$ is the union of two
  %patterns that are order-isomorphic to $\sigma$.
  Let $2n = |\pi|$ (and hence $|\sigma| = n$).
  %Fix the occurrences in $\pi$ of the two disjoint patterns that
  %are order-isomorphic to $\sigma$ and
  %write
  %$I^1 = \{i^1_1, i^1_2, \ldots, i^1_n\}$,
  %$i^1_1 < i^1_2 < \ldots < i^1_n$, for the positions in $\pi$
  %of the first pattern order-ismorphic to $\sigma$ and
  %$I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$,
  %$i^2_1 < i^2_2 < \ldots < i^2_n$, for the positions in $\pi$
  %of the second pattern order-ismorphic to $\sigma$.
  Let $G_\pi$ be the linear clique associated to $\pi$ with
  vertices $\VS{G_\pi} =
  \{i^1_j : 1 \leq j \leq n\} \cup \{i^2_j : 1 \leq j \leq n\}$.
  It is easily seen that
  $\mathcal{M} = \{\{i^1_j, i^2_j\} : 1 \leq j \leq n\}$
  is a subset of $\ES{G_\pi}$.
  Furthermore, $\mathcal{M}$ is a perfect matching since
  $I^1 \cap I^2 = \emptyset$,
  and $I^1 \cup I^2 = \{1, 2,  \ldots 2n\}$.
  It is also containment-free.
  Indeed, if it is not the case then there would exist two edges
  $e = \{i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$, $j < k$,
  in $\mathcal{M}$ such that
  $i^1_j < i^1_k < i^2_k < i^2_j$ or
  $i^1_j < i^2_k < i^1_k < i^2_j$.
  This is a contradiction since
  $i^2_k > i^2_j$ if $k > j$.
  Finally, for any two distinct edges
  $e = \{\i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
  we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
  $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing
  in both cases two elements (at position $j$ and $k$)
  in two patterns that are order-isomorphic to $\sigma$.

@@

\begin{proof}
  The problem is certainly in $\NPclass$.
  We propose a reduction from the pattern involvement problem which is
  known to be \NPC~\cite{Bose:Buss:Lubiw:1998}:
  Given two permutations $\pi$ and $\sigma$, decide whether $\sigma$
  occurs in $\pi$ (as an order-isomorphic pattern).

@@

      \node (claim edge 1) at (8.5,8.5) {Claim~\ref{claim:(nu_4, nu'_4)-edge}};

@@

      % \node (claim edge 1) at (6.5,4) {Claim~\ref{}};

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?)
% Given two
% permutations $\pi$ and $\sigma$, how hard is the problem of deciding
% whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip
